### 2025-11-08 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
2025NSF: better flow and removed repetitions

Before looking at our preliminary proposal for the axiomatic, let us develop some intuition about the structure we need to generalize. Note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, the biggest problem is understanding how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we have found the right approach. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given, in the classical case, when the two measures have disjoint support; in the quantum case, it is given when the eigenspaces of non-zero eigenvalues are orthogonal. Our approach, then, is to proceed in the opposite way: impose a generic entropy function to have an upper bound and enough properties to show that said upper bound must correspond to the Shannon entropy formula~\cite{Shannon}. Since both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case as a recursive expansion of the upper bound.

Another important insight from our preliminary work is that the entropy is not only responsible for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. The idea, then, is to recover to the geometric structure of the state space from it entropic structure.

@@

Given that \emph{all} physical theories must admit a notion of ensembles that can be statistically mixed via convex combination, these insights from our preliminary work show that the entropic structure plays a crucial role in all physical theories and merits further study. Moreover, while typically information geometry is cast in finite-dimensional spaces, there is nothing preventing its generalization to infinite dimensions. If the entropic structure, in fact, imposes a distance function whose open balls are convex, then this provides us with a locally convex topology which is exactly what is needed to define calculus in infinite dimensions. Moreover, the Lie algebras of both classical and quantum mechanics, Poisson brackets and commutators, respectively, have a special relationship with the geometric and entropic structures, which may provide us with a template to generalize those structures as well.

Given these very promising preliminary findings, the goal of the proposed research is to find a minimal set of axioms to characterize the entropic structure of physical theories. Minimal in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics~\cite{RevMathBook}, the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Let us, then, go through the current state of our preliminary work, and show the type of questions that the overall approach touches.

>>>>

Before proceeding, we should not that the entropic structure is not crucial for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. One direction we want to pursue, in fact, is to recover to the geometric structure of the state space from it entropic structure.

@@

Lastly, as port of our larger program, we have shown that classical mechanics emerges as the high entropy limit of quantum mechanics. These insights show that the entropic structure is more fundamental that one may expect, and that finding its axiomatic characterization should allow for a more precise generalization of a series of results. We stress that the axioms that characterize the entropic structure of physical theories should be minimal, in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics~\cite{RevMathBook}, the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Part of the work, then, is to understand the implication relationships between different requirements. Lastly, note that the Lie algebras of both classical and quantum mechanics, Poisson brackets and commutators, respectively, have a special relationship with the geometric and entropic structures, which may provide us with a template to generalize those structures as well.

Before delving into the axiomatic, note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, we need to understand how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we have found the right approach. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given, in the classical case, when the two measures have disjoint support; in the quantum case, it is given when the eigenspaces of non-zero eigenvalues are orthogonal. Our approach, then, is to proceed in the opposite way: impose a generic entropy function to have an upper bound and enough properties to show that said upper bound must correspond to the Shannon entropy formula~\cite{Shannon}. Since both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case as a recursive expansion of the upper bound.

Having justified why studying the entropic structure of ensemble spaces is of interest and having found a fruitful approach, let us go through the axioms of each structure from our preliminary work.
